---
title: "Introduction to tukeyAdEnet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tukeyAdEnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5
)
```

## Motivation

Classical penalised regression estimators such as the lasso, elastic net,
and their adaptive variants minimise a squared-error loss. Squared error is
unbounded, so a single large outlier in the response, or a single leverage
point in the design matrix, can dominate the objective and badly distort
the fitted coefficients.

**tukeyAdEnet** replaces the squared-error loss with Tukey's redescending
biweight loss. Its derivative (the influence function) is exactly zero for
standardised residuals beyond a tuning constant `d`, so observations far
enough from the bulk of the data have *no* influence on the fit, rather
than merely a bounded one (as with, say, Huber loss). Combined with an
adaptive elastic net penalty, this gives a sparse estimator with a positive
breakdown point.

```{r}
library(tukeyAdEnet)

u <- seq(-8, 8, length.out = 401)
op <- par(mfrow = c(1, 2))
plot(u, tukeyRho(u), type = "l", main = "Biweight loss", ylab = expression(rho(u)))
plot(u, tukeyPsi(u), type = "l", main = "Influence function", ylab = expression(psi(u)))
abline(v = c(-4.685, 4.685), lty = 2, col = "grey50")
par(op)
```

## Fitting a single model

`tukeyAdEnet()` fits the model at a given `(lambda1, lambda2)` pair. It
returns adaptively weighted coefficients, using either a robust
(`robustbase::lmrob()`, when available and `p < n`) or ridge initial fit to
build the adaptive weights `w_j = 1 / |beta_init_j|`.

```{r}
set.seed(1)
n <- 150
p <- 12
x <- matrix(rnorm(n * p), n, p)
beta_true <- c(3, -3, 2, rep(0, p - 3))
y <- as.numeric(x %*% beta_true + rnorm(n))

# contaminate 10% of the responses with gross outliers
contaminated <- sample(seq_len(n), size = round(0.1 * n))
y[contaminated] <- y[contaminated] + rnorm(length(contaminated), mean = 30, sd = 5)

fit <- tukeyAdEnet(x, y, lambda1 = 0.3, lambda2 = 0.05)
print(fit)
coef(fit)
```

For comparison, an ordinary least-squares fit on the same contaminated data
is badly biased by the 10% outliers:

```{r}
ols <- coef(lm(y ~ x - 1))
data.frame(
  truth = beta_true,
  tukeyAdEnet = as.numeric(fit$beta),
  ols = as.numeric(ols)
)
```

## Automatic tuning via robust BIC

Cross-validation is itself sensitive to outliers landing in a held-out
fold. `tukeyAdEnetRBIC()` instead searches a two-dimensional
`(lambda1, lambda2)` grid and selects the pair minimising a robust BIC
criterion (the biweight loss plus a `log(n)` penalty on the number of
non-zero coefficients).

```{r}
tuned <- tukeyAdEnetRBIC(x, y, n_lambda1 = 15, lambda2_factors = c(0, 0.05, 0.2, 0.5))
tuned
coef(tuned)
```

```{r, fig.alt = "Robust BIC against lambda1, one curve per lambda2 value"}
plot(tuned)
```

## Prediction

Both `tukeyAdEnet` and `tukeyAdEnetRBIC` objects support `predict()` on new
data:

```{r}
x_new <- matrix(rnorm(5 * p), 5, p)
predict(tuned, newx = x_new)
```

## Notes on scaling

`x` should not include an intercept column, and the elastic net penalty is
not scale-invariant, so standardising the columns of `x` (e.g. with
`scale()`) before fitting is recommended when predictors are not already
on comparable scales.
